Inferring probabilistic stellar rotation periods using Gaussian processes

### 2017-02-02 documents/ms.tex
fixing double Suzanne in author list

\author{%
   Ruth Angus\altaffilmark{\columbia},
   Timothy Morton\altaffilmark{\princeton}
   Suzanne Aigrain\altaffilmark{\oxford},
   Daniel Foreman-Mackey\altaffilmark{\uw}
   Suzanne Aigrain\altaffilmark{\oxford},
   \& Vinesh Rajpaul\altaffilmark{\oxford}
}

@@

The light curves of spotted, rotating stars are often non-sinusoidal and
Quasi-Periodic (QP).
These stars vary in brightness due to active regions on their surfaces which
rotate in and out of view.
Complicated surface spot patterns produce the non-sinusoidal quality,
and the finite lifetimes of these active regions and the presence of 
differential rotation on the stellar surface produce quasi-periodicity.
A strictly periodic sinusoid therefore does not well describe 
rotationally modulated stellar light curves.
A physically realistic model of the stellar surface 
would perfectly capture the complexity of shapes
within stellar light curves as well as the quasi-periodic nature, allowing for
extremely precise probabilistic period recovery when conditioned on the data.
However, such physical models require many free parameters in order to
accurately represent a stellar surface and some of these parameters are
extremely degenerate \citep[\eg][]{Russell1906, Jeffers2009, Kipping2012}.
In addition to global stellar parameters such as inclination and rotation
period, each spot or active region should have (at minimum) a longitude,
latitude, size, temperature and lifetime.
Considering that many stars likely have hundreds of spots, the number
of free parameters quickly becomes unwieldy, especially to explore
their posterior PDFs with MCMC.
Simplified spot models, such as the one described in \citet{Lanza2014} where
only two spots are modelled, have produced successful results; however, such 
relatively inflexible models sacrifice precision.

The standard methods used for measuring rotation periods include detecting
peaks in a Lomb-Scargle \citep{Lomb1976, Scargle1982} (LS) periodogram
\citep[e.g.][]{Reinhold2013}, Auto-Correlation Functions (ACFs),
\citep{Mcquillan2013} and wavelet transforms \citep{Garcia2014}.
The precision of the LS periodogram and wavelet methods are limited by the
suitability of the model choice: a sinusoid for the LS periodogram, 
and a choice of mother wavelet, assumed to desribe the data
over a range of transpositions \citep[see, \eg][]{Carter2010}, 
for the wavelet method.
In contrast, the ACF method is much better suited to signals that are
non-sinusoidal.
In fact, as long as it is approximately periodic the ACF will 
display a peak at the rotation period, no matter the signal shape.
A drawback of the ACF method, however, is that it requires data to be
evenly-spaced\footnote{\citet{Edelson1988} describe a method for computing
ACFs for unevenly-spaced data.}, which is not exactly the case with \Kepler\
light curves (although in many cases it can be approximated as uniformly
sampled) and not nearly the case for the future large-scale survey instrument
\LSST.
An ACF is also an operation performed on the data, not a generative model of
the data, and so not inherently probabilistic.
Therefore, there is no justified way to estimate the uncertainty on an ACF
rotation period measurement.
% Many rotation periods in the literature have been inferred by measuring the
% position of the first peak in an ACF, however this approach can be dangerous.
% The exponential decay in correlation can shift the peak position short-wards
% of its true value, leading to an underestimate of the period.
% We return to this point in section \textsection \ref{sec:discussion}.

In this work, we introduce an {\it effective} model for rotationally
modulated stellar light curves
which captures the salient behaviour but is not
physically motivated---although some parameters may indeed be 
{\it interpreted} as physical ones.
An ideal effective model should have a small number of non-degenerate parameters 
and be flexible enough to perfectly capture non-sinusoidal, QP behaviour.
A Gaussian process (GP) model fulfills these requirements. We use this GP 
model as the generative model at the backbone of a method to probabilistically 
infer accurate and precise stellar rotation periods.  This enables us to 
estimate the posterior PDF of the rotation period, thereby producing a 
justified estimate of its uncertainty while simultaneously marginalizing over
parameters which model correlated noise.

GPs are commonly used in the machine learning community and increasingly
in other scientific fields, for example biology and geophysics.
More recently, GPs have been used in the astronomy literature \citep[see
\eg][]{Gibson2012, Haywood2014, Haywood2015, Evans2015, Rajpaul2015,
Rajpaul2016, Aigrain2016}.
They are useful in regression problems involving any stochastic process,
specifically when the probability distribution for the process is a
multi-variate Gaussian.
If the probability of obtaining a dataset is a Gaussian in $N$ dimensions,
where $N$ is the number of data points, a GP can describe that dataset. 
An in-depth introduction to Gaussian processes is provided by
\citet{Rasmussen2005}.

GP models parameterise the covariance between data points by means of a kernel
function that defines the form of the covariance matrix.
For example, take the time-series in Figure \ref{fig:GP_example}: 
the \kepler\ light curve of KIC \kepexample, a relatively active star that
rotates once every $\sim$ 4 days, with stochastic variability typical of 
\kepler\ FGK stars.
Clearly, data points in this light curve are correlated.
Points close together in time are tightly correlated and points more
widely separated are loosely correlated.
A GP models the way in which this correlation varies with the separation between 
data points; that is, it models not the data directly but rather the {\it 
covariance structure}.
This gives GPs their flexibility---they can model any time
series with a similar covariance structure.
In addition, a very simple function can usually capture the covariance
structure of a light curve, whereas a much more complicated one may be
required to model the time series itself.
Figure \ref{fig:GP_example} demonstrates how the light curve of KIC \kepexample\ 
may be modelled with a GP.
% Both provide adequate fits to the data, however only the periodic kernel
% function, `QP' is a useful model because it has a periodic parameter.
% I return to this point shortly.

A range of kernel functions can describe stellar variability.
For example, the simplest and most commonly used kernel function, the `Squared
Exponential' (SE), defined as follows, 
produces an adequate fit to the KIC \kepexample\ light curve:
\begin{equation}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2} \right).
\end{equation}
\label{eq:SE}
Here $A$ is the amplitude of covariance, $l$ is the length scale of
exponential decay, and $x_i-x_j$ is the separation between data points.
The SE kernel function has the advantage of being very simple, with
just two parameters, $A$ and $l$.
If $l$ is large, two data points far apart in $x$ will be tightly correlated,
and if small they will be loosely correlated.
Another property of the SE kernel function is that it produces functions that
are infinitely differentiable.
It is therefore possible to model a data set and its derivatives
simultaneously.
% \citep[see \eg][]{Rajpaul2015}

However, The SE kernel function is not a good model of the covariance 
in stellar light curves, nor a {\it useful} one for the problem of 
rotation period inference because it does not capture periodic behaviour.
In order to infer rotation periods it is thus necessary to use a periodic kernel
function.  For this reason, we use the `Quasi-Periodic' kernel, defined as
\begin{equation}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2} -
\Gamma^2 \sin^2(\frac{2\pi}{P}) \right).
\end{equation}
\label{eq:QP}
It is the product of the SE kernel function, which describes the overall
covariance decay, and an exponentiated, squared, sinusoidal kernel function
that describes the periodic covariance structure.
$P$ can be interpreted as the rotation period of the star and $\Gamma$
controls the amplitude of the $\sin^2$ term.
If $\Gamma$ is very small, only points almost exactly one period away are
tightly correlated and points that are slightly more or less than one period
away are very loosely correlated.
If $\Gamma$ is large, points separated by one period are tightly
correlated and points separated by slightly more or less are still highly
correlated although less so.
This kernel function allows two data points that are separated in time by one
rotation period to be tightly correlated, while also allowing 
points separated by half a period to be weakly correlated.
We also use an extra parameter $\sigma$ which is an additional white noise
term added to the diagonal elements of the covariance matrix.
It is the fraction by which the observational uncertainties have been
underestimated --- if the uncertainties reported on the data are too small,
this parameter will be non-zero.
In practice however, this parameter should always be non-zero when performing
GP regression.
This is because the covariance matrix must be positive definite; however,
matrix inversion performed using most solvers is approximate, not exact,
so slight deviations from positive definiteness can arise.
Including a small amount of extra variance in the model allows enough
flexibility that the matrix inversion algorithms do not run into numerical
issues.  We use this QP kernel function to produce the GP model 
that fits the \Kepler\ light curve in in Figure \ref{fig:GP_example}.

To infer a stellar rotation period from a light curve, we fit this QP-kernel
GP model to the data.  As with any model-fitting exercise, the likelihood
of the model could be maximized to find the maximum-likelihood value for $P$.
In this study, however, we explore the full posterior PDFs using MCMC.  While
this approach comes at a computational cost, such posterior exploration 
importantly provides a justified estimate of the uncertainty on rotation period.

@@

In \textsection \ref{sec:simulations} of this paper we describe the simulated
light curves used to test the three rotation period recovery methods---periodogram,
ACF, and GP---that we describe in detail in \textsection \ref{sec:method}.
In \textsection \ref{sec:kepler} we apply our method to real \kepler\ data, and
we discuss the results in \textsection \ref{sec:discussion}.

>>>>

\author{%
   Ruth Angus\altaffilmark{\columbia},
   Timothy Morton\altaffilmark{\princeton},
   Suzanne Aigrain\altaffilmark{\oxford},
   Daniel Foreman-Mackey\altaffilmark{\uw}
   \& Vinesh Rajpaul\altaffilmark{\oxford}
}

@@

The light curves of spotted, rotating stars are often non-sinusoidal and
Quasi-Periodic (QP).
These stars vary in brightness due to active regions on their surfaces which
rotate in and out of view.
The non-sinusoidal quality is caused by the complicated surface spot patterns
and the quasi-periodicity is produced both by the finite lifetimes of these
active regions and the presence of differential rotation on the stellar
surface.
A strictly periodic sinusoid is therefore not a good model for stellar light
curves.
In an ideal world, a physical model of the stellar surface would be
conditioned on the data.
A physically realistic model would perfectly capture the complexity of shapes
within stellar light curves as well as the quasi-periodic nature, allowing for
extremely precise probabilistic period recovery.
However, such physical models require many free parameters in order to
accurately represent a stellar surface and some of these parameters are
extremely degenerate \citep[\eg][]{Russell1906, Jeffers2009}.
In addition to global stellar parameters such as inclination and rotation
period, each spot or active region should have (at minimum) a longitude,
latitude, size, temperature and lifetime.
Considering that many stars have on the order of hundreds of spots, the number
of free parameters quickly becomes unwieldy, especially if the posterior PDFs
of these parameters are explored with MCMC.
Simplified spot models, such as the one described in \citet{Lanza2014} where
only two spots are modelled, have produced successful results, however these
simplified models sacrifice some precision due to lack of model flexibility.
Instead of using a physical model for stellar light curves, we choose to use
an {\it effective} model: one which captures the behaviour but is not
physically motivated, although the parameters of this model may be {\it
interpreted} as physical ones.
An ideal effective model for the light curve of a spotted, rotating star is
one with a small number of non-degenerate parameters that is flexible enough
to perfectly capture non-sinusoidal, QP behaviour.
These requirements are fulfilled by a Gaussian process (GP) model.

The standard methods used for measuring rotation periods include detecting
peaks in a Lomb-Scargle \citep{Lomb1976, Scargle1982} (LS) periodogram
\citep[e.g.][]{Reinhold2013}, Auto-Correlation Functions (ACFs)
\citep{Mcquillan2013} and wavelet transforms \citep{Garcia2014}.
The precision of the LS periodogram and wavelet methods are limited by the
suitability of the model choice.
A sinusoid is used in the case of the LS periodogram and the wavelet method
relies on a choice of mother wavelet that is assumed to describe the data over
a range of transpositions \citep[see, \eg][]{Carter2010}.
In contrast, the ACF method is much better suited to signals that are
non-sinusoidal.
In fact, it doesn't matter what shape the signal is: as long as it is
approximately periodic the ACF will display peaks located at the rotation
period.
A drawback of the ACF method however, is that it requires data to be
evenly-spaced\footnote{\citet{Edelson1988} describe a method for computing
ACFs for unevenly-spaced data.} which is not exactly the case with \Kepler\
light curves (although in many cases it can be approximated as uniformly
sampled) and not nearly the case for the future large-scale survey instrument,
\LSST.
An ACF is also an operation performed on the data, not a generative model of
the data and is therefore not inherently probabilistic.
For this reason it is very difficult to estimate the uncertainty on an ACF
rotation period measurement.
% Many rotation periods in the literature have been inferred by measuring the
% position of the first peak in an ACF, however this approach can be dangerous.
% The exponential decay in correlation can shift the peak position short-wards
% of its true value, leading to an underestimate of the period.
% We return to this point in section \textsection \ref{sec:discussion}.

The motivation for developing a GP-based method for rotation period inference
is, firstly to measure more accurate and precise rotation periods using a
better-suited generative model than a sinusoid for the reasons explained
above.
Secondly, in order to infer {\it probabilistic} periods, i.e. to estimate the
posterior PDF of the rotation period and thereby produce a realistic and
estimate of its uncertainty.
Thirdly, to allow for an additional correlated noise model to be included
during regression, the parameters of which could be marginalised over.
% And fourthly to provide some way of determining whether a periodic model is
% supported by the data over a purely stochastic one.

GPs are commonly used in the machine learning community and increasingly
in other scientific fields, for example biology and geophysics.
More recently, GPs have been used in the astronomy literature \citep[see
\eg][]{Gibson2012, Haywood2014, Haywood2015, Evans2015, Rajpaul2015,
Rajpaul2016, Aigrain2016}.
% FIXME: CITATIONS Gibson, Aigrain, Evans, Rajpaul, Waldman(?), Cosmology,
% geology, biology, Rasmussen & Williams, etc.
They are useful in regression problems involving any stochastic process,
specifically when the probability distribution for the process is a
multi-variate Gaussian.
If the probability of obtaining a data-set is a Gaussian in $N$ dimensions,
where $N$ is the number of data points, that data-set can be described as, and
with, a Gaussian process.
An in-depth introduction to Gaussian processes is provided by
\citet{Rasmussen2005}.

GP models parameterise the covariance between data points and a kernel
function provides the form of covariance matrix parameterisation.
For example, take the time-series in figure \ref{fig:GP_example}.
This is the \kepler\ light curve of KIC \kepexample, a relatively active star that
rotates once every $\sim$ 4 days.
The stochastic variability in this time-series is typical of \kepler\ FGK
stars.
Clearly, data points in this light curve are correlated.
Points that are close together in time are tightly correlated and those that
are widely separated in time are loosely correlated.
The way in which the covariance varies with the separation between data points
is modelled when using GP regression; it is not the data but the {\it
covariance structure} that is modelled.
This fact is what gives GPs their flexibility---they can model any time
series with a similar covariance structure.
In addition, a very simple function can usually capture the covariance
structure of a light curve, whereas a much more complicated one may be
required to model the time-series itself.
% The light curve of \Kepler-452 has been modelled with a GP in figure
% \ref{fig:GP_example}, shown in orange.
In figure \ref{fig:GP_example} the light curve of KIC \kepexample\ has been
modelled with a GP.
% Both provide adequate fits to the data, however only the periodic kernel
% function, `QP' is a useful model because it has a periodic parameter.
% I return to this point shortly.

A range of kernel functions could be used to describe stellar variability.
For example, the simplest and most commonly used kernel function, the `Squared
Exponential' (SE) produces an adequate fit to the KIC \kepexample\ light
curve.
The SE kernel function is defined as,
\begin{equation}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2} \right).
\end{equation}
\label{eq:SE}
Here $A$ is the amplitude of covariance, $l$ is the length scale of
exponential decay and $x_i-x_j$ is the separation between data points.
The SE kernel function has the advantage of being very simple---it has just
two parameters, a covariance amplitude and length scale: $A$ and $l$.
If $l$ is large two data points far apart in $x$ will be tightly correlated,
and if small they will be loosely correlated.
Another property of the SE kernel function is that it produces functions that
are infinitely differentiable.
It is therefore possible to model a data set and its derivatives
simultaneously.
% \citep[see \eg][]{Rajpaul2015}
The SE kernel function is not a good model of the covariance in stellar light
curves, nor a {\it useful} one for the problem of rotation period inference
because it does not capture periodic behaviour.
In order to infer rotation periods it is necessary to use a periodic kernel
function.
For this reason, we use the `Quasi-Periodic' kernel function.
The QP kernel function is defined as
\begin{equation}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2} -
\Gamma^2 \sin^2(\frac{2\pi}{P}) \right).
\end{equation}
\label{eq:QP}
It is the product of the SE kernel function, which describes the overall
covariance decay, and an exponentiated, squared, sinusoidal kernel function
that describes the periodic covariance structure.
$P$ can be interpreted as the rotation period of the star and $\Gamma$
controls the amplitude of the $\sin^2$ term.
If $\Gamma$ is very small, only points almost exactly one period away are
tightly correlated and points that are slightly more or less than one period
away are very loosely correlated.
If $\Gamma$ is large, points separated by one period are tightly
correlated and points separated by slightly more or less are still highly
correlated although less so.
This kernel function allows two data points that are separated in time by one
rotation period to be tightly correlated, while points separated by half a
period can be weakly correlated.
We also use an extra parameter, $\sigma$ which is an additional white noise
term added to the diagonal elements of the covariance matrix.
It is the fraction by which the observational uncertainties have been
underestimated --- if the uncertainties reported on the data are too small,
this parameter will be non-zero.
In practice however, this parameter should always be non-zero when performing
GP regression.
This is because the covariance matrix must be positive definite, however
matrix inversion performed using most solvers is approximate, not exact,
therefore slight deviations from positive definiteness can arise.
Including a small amount of extra variance in the model allows enough
flexibility that the matrix inversion algorithms do not run into numerical
issues.

This QP kernel function was used to produce the model shown in figure
~\ref{fig:GP_example}.

In order to infer a stellar rotation period from a light curve, we fit a GP
model with a QP kernel function to the data.
The likelihood of the model, conditioned on the data could then be maximised
in order to find the maximum likelihood value for $P$.
In this study however, the full posterior PDFs of the parameters are explored
using MCMC.
This approach comes at a cost: a GP model is expensive to compute once as it
requires an inversion and determinant evaluation of the covariance matrix, let
alone the many thousands of times that is necessary to fully explore the
posteriors of the parameters.
Fully exploring the posterior PDF of $P$ is important however, as it provides
an accurate estimate of the uncertainty on the rotation period.

@@

In \textsection \ref{sec:simulations} of this paper we describe the simulated
light curves used to test rotation period inference methods.
In \textsection \ref{sec:method} we describe the various methods used to
recover the rotation periods: the periodogram, ACF and GP methods.
In \textsection \ref{sec:kepler} we apply our method to real \kepler\ data and
the results of the paper are discussed in \textsection \ref{sec:discussion}.
